How important are functional and developmental constraints on phenotypic evolution? An empirical test with the stomatal anatomy of flowering plants

### 2022-11-01 ms/ms.tex
Moved online supplement to separate document for publication

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% \documentclass[11pt]{article} % redundant
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\usepackage{multirow} % for merging rows in tables

@@

\setstretch{1.7}
\begin{flushleft}
{\Large
\textbf\newline{How important are functional and developmental constraints on phenotypic evolution? An empirical test with the stomatal anatomy of flowering plants}
}
\newline
\\
Christopher D. Muir\textsuperscript{1,*},
Miquel \`{A}ngel Conesa\textsuperscript{2},
Jeroni Galm\'{e}s\textsuperscript{2},
Varsha S. Pathare\textsuperscript{3},
Patricia Rivera\textsuperscript{4},
Rosana López Rodríguez\textsuperscript{5},
Teresa Terrazas\textsuperscript{4},
Dongliang Xiong\textsuperscript{6}
\\
\bigskip
\bf{1} School of Life Sciences, University of Hawaii at M\=anoa, Honolulu, HI 96822, USA \\
\bf{2} Research Group on Plant Biology under Mediterranean Conditions, Departament de Biologia, Universitat de les Illes Balears, Ctra. Valldemossa km 7.5, E-07122, Palma, Spain \\
\bf{3} School of Biological Sciences, Washington State University, Pullman, WA 99164-4236, USA \\
\bf{4} Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Apartado Postal 70‑367, 04510 Mexico City, Mexico \\
\bf{5} Departamento de Sistemas y Recursos Naturales, Universidad Politécnica de Madrid, 28040 Madrid, Spain \\
\bf{6} National Key Laboratory of Crop Genetic Improvement, MOA Key Laboratory of Crop Ecophysiology and Farming System in the Middle Reaches of the Yangtze River, College of Plant Science and Technology, Huazhong Agricultural University, Wuhan, Hubei 430070, China\\
\bigskip
* cdmuir@hawaii.edu

@@

\linenumbers{}
\modulolinenumbers[3]

@@

Given the significance of stomata for plant function and global vegetation modeling (\protect\hyperlink{ref-berry_stomata:_2010}{Berry, Beerling, and Franks 2010}), we would like to understand what factors constrain their anatomical variation. Here we focus on interspecific variation in mean trait values rather than intraspecific variation. Packing constraints and developmental integration could explain inverse size-density scaling. The density and size of a stomata on a planar leaf surface can be viewed as a packing problem where the total area allocated to stomata cannot exceed the total leaf area. This is a functional, or geometric, constraint because certain combinations of large size and high density are not physically possible. This functional constraint cannot explain why combinations of low density and small size are rare, but may explain why stomatal size must decrease when density increases as the leaf runs out of space. The packing limit of functional stomata is less than the entire leaf area, but the exact value is unclear. The realized upper limit is close to 1/3 or 1/2 (\protect\hyperlink{ref-de_boer_optimal_2016}{de Boer et al. 2016}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}; \protect\hyperlink{ref-liu_scaling_2021}{Liu et al. 2021}) for the species' mean, not an individual leaf.

@@

The second hypothesis is the cell size mediates developmental integration between stomatal size and density. If developmental integration is the primary reason for inverse size-density scaling, then amphistomatous leaves will exhibit identical size-density scaling on each surface. If the stomatal size and density scale differently on each surface, this implies that they can evolve independently and that selection against generally maladaptive trait combinations shapes their covariance. Furthermore, we predicted that divergence in genome size, which is strongly associated with meristematic cell volume (\protect\hyperlink{ref-simova_geometrical_2012}{Šímová and Herben 2012}), would covary with stomatal size and density similarly on each surface.

@@

Unless otherwise mentioned, we performed all data wrangling and statistical analyses in \emph{R} version 4.2.0 (\protect\hyperlink{ref-r_core_team_r:_2022}{R Core Team 2022}). Source code is publicly available on GitHub (\url{https://github.com/cdmuir/stomata-independence}) and will be archived on Zenodo upon publication.

@@

Developmental integration could shape patterns of phenotypic macroevolution, but a major hindrance to progress is that verbal models do not make precise, quantitative predictions that distinguish it from alternatives. An advantage of testing developmental integration in stomata is that their development is well studied (\protect\hyperlink{ref-bergmann_stomatal_2007}{Bergmann and Sack 2007}; \protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}). We can leverage that knowledge to build a developmental function and derive equations for phenotypic (co)variance caused by developmental integration. If observed patterns of evolution are inconsistent with developmental integration, theory may also help identify which parameters lead to developmental disintegration. We combined and extended stomatal development models to predict how stomatal density and length would diverge if stomatal development were constrained and how those predictions would change if stomatal development were unconstrained. We summarize our methods verbally here and direct readers interested in the mathematical details to Notes A1. A graphical summary is provided in Fig. \ref{fig:developmental-integration}. We imposed constraint by assuming the stomatal developmental function is constrained. The developmental function maps cell size prior to differentiation onto stomatal size and density using two parameters. The first parameter describes how cell volume is apportioned between epidermal cells and guard cell meristemoids during asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}). The second parameter is stomatal index (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}), which is determined by amplifying and spacing divisions after asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}). When these parameters are fixed, divergence in stomatal size and density is determined by divergence in meristematic cell volume and expansion prior to asymmetric division. We relaxed this constraint by treating parameters of the developmental function as random variables that can diverge between species. We used random variable algebra to derive predicted (co)variance in divergence between stomatal length and density (see \protect\hyperlink{ref-lynch_genetics_1998}{Lynch and Walsh 1998} for key random variable algebra theorems).

@@

We searched the literature for studies that measured stomatal density and stomatal size, either guard cell length or stomatal pore length, for both abaxial and adaxial leaf surfaces. In other words, we did not include studies unless they reported separate density and size values for each surface. We did not record leaf angle because it is typically not reported, but we presume that for the vast majority of taxa that the abaxial is the lower surface and the adaxial is the upper surface. This is reversed in resupinate leaves, but to the best of our knowledge, our synthesis did not include resupinate leaves. None of the species with resupinate leaves listed by Chitwood et al. (\protect\hyperlink{ref-chitwood_conflict_2012}{2012}) are in our data set. We refer to guard cell length as stomatal length and converted stomatal pore length to stomatal length assuming guard cell length is twice pore length (\protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}). Table \ref{tab:traits} lists focal traits and symbols.

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2000}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5176}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2824}}@{}}
\caption{\label{tab:traits}Stomatal anatomical traits with mathemtical symbol, variable string used in source code, and scientific units.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\centering
Symbol
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Variable string
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Units
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\centering
Symbol
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Variable string
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Units
\end{minipage} \\
\midrule()
\endhead
\(D_\mathrm{ab}\) & \(\mathtt{abaxial\_stomatal\_density\_mm2}\) & \(\text{pores mm}^{-2}\) \\
\(D_\mathrm{ad}\) & \(\mathtt{adaxial\_stomatal\_density\_mm2}\) & \(\text{pores mm}^{-2}\) \\
\(L_\mathrm{ab}\) & \(\mathtt{abaxial\_stomatal\_length\_um}\) & \(\mu\)m \\
\(L_\mathrm{ad}\) & \(\mathtt{adaxial\_stomatal\_length\_um}\) & \(\mu\)m \\
\bottomrule()
\end{longtable}

@@

We generated an ultrametric, bifurcating phylogeny of 638 taxa by resolving and removing ambiguous taxonomic names, placing taxa on the GBOTB.extended mega-tree of seed plants (\protect\hyperlink{ref-smith_constructing_2018}{S. A. Smith and Brown 2018}; \protect\hyperlink{ref-zanne_three_2014}{Zanne et al. 2014}), and resolving polytomies using published sequence data. Divergence times in this phylogeny are based on extensive fossil calibration (see \protect\hyperlink{ref-smith_constructing_2018}{S. A. Smith and Brown 2018}; \protect\hyperlink{ref-magallon_metacalibrated_2015}{Magallón et al. 2015} for further detail). The complete methodology is described in the online supplement (Notes A2).

@@

All hypotheses make predictions about trait (co)variance matrices or parameters derived from them (see Notes A1 and subsections below). Within and among species covariation is a hallmark of developmental integration (\protect\hyperlink{ref-armbruster_multilevel_1988}{Armbruster 1988}), but other evolutionary processes also lead to covariance. Distinguishing between them requires deriving predictions and testing whether observed covariance is consistent with one hypothesis or another. We estimated the \(4 \times 4\) covariance matrix of phylogenetically independent contrasts between log-transformed values of \(\Delta \text{log}(D_\mathrm{ab})\), \(\Delta \text{log}(D_\mathrm{ad})\), \(\Delta \text{log}(L_\mathrm{ab})\), and \(\Delta \text{log}(L_\mathrm{ad})\) using a distributional multiresponse robust Bayesian approach. See Table \ref{tab:traits} for variable definitions. We denote variances as \(\text{Var}[\Delta \text{log(trait)}]\) and covariances as \(\text{Cov}[\Delta \text{log(trait}_1),\Delta \text{log(trait}_2)]\). We used a multivariate \(t\)-distribution rather than a Normal distribution because estimates using the former are more robust to exceptional trait values (\protect\hyperlink{ref-lange_robust_1989}{Lange, Little, and Taylor 1989}). We also estimated whether the variance in trait divergence increases with time. Under many trait evolution models (e.g.~Brownian motion), interspecific variance increases through time. To account for this, we included time since taxon-pair divergence as an explanatory variable affecting the trait covariance matrix.

@@

We asked how divergence in stomatal length and density would covary if the developmental function were constrained and compared it to their divergence when the developmental function can evolve. When the developmental function is constrained this means that allocation to guard cell meristemoids during asymmetric division and stomatal index are fixed (see Notes A1 for mathematical description). Under these assumptions, divergence in stomatal length and density is mediated entirely by divergence in meristematic cell volume and expansion prior to differentiation. Developmental integration is strong because divergence in density is perfectly negatively correlated with divergence in size. In contrast, stomatal length and density can diverge independently when the developmental function is not fixed. Divergence in asymmetric cell division affects stomatal size independently of density; divergence in stomatal index affects stomatal density independently of size. Divergence in the developmental function causes developmental disintegration because stomatal density and size can diverge independently. Developmental integration is minimal when asymmetric cell division and/or stomatal index diverge more than meristematic cell volume and expansion. The three main conclusions are that 1) developmental constraint leads to developmental integration; 2) different (co)variance in divergence of stomatal length and density on each surface implies the developmental function is not fixed; and 3) divergence in different components of the developmental function affect stomatal length and density differently. See Notes A1 and Table \ref{tab:predictions} for more a complete derivation and detailed predictions.

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/h3.pdf}
\caption{Evolutionary divergence slows down as epidermal space fills up. This pattern is consistent with functional constraints (packing limits) constraining evolution, but only when stomata occupy a large fraction of epidermal area. The shaded area in each facet indicates the estimate of where 95\% of the 236 phylogenetically independent contrasts fall as a function of the fraction of epidermal area allocated to stomata per surface. Each point is the absolute value of  $\Delta~\textrm{log(trait)}$ for stomatal density (left facets) or length (right facets) on the adaxial (upper facets) and abaxial (lower facets) surface. The fraction of epidermal area allocated to stomata is the average value per surface between the two taxa in each contrast. Divergence in anatomical traits is more variable when stomata occupy a smaller area, especially for adaxial stomatal density (upper left facet). See Table \ref{tab:h3output} for all parameter estimates and confidence intervals.}
\label{fig:h3}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/h1-raw.pdf}
\caption{Inverse size-density scaling differs between the surfaces of amphistomatous leaf, indicating different phenotypic constraints. The panels show the relationship between stomatal length ($x$-axis) and stomatal density ($y$-axis) on a log-log scale for values measured on the abaxial leaf surface (left) and the adaxial leaf surface (right) across 638 taxa.}
\label{fig:h1_raw}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/h1.pdf}
  \caption{(Caption next page.)}
  \label{fig:h1}
\end{figure}
\addtocounter{figure}{-1}

\begin{figure} [t!]
\caption{(Previous page.) Greater overall phenotypic constraint on abaxial stomatal anatomy is evinced by more variable evolutionary divergence in adaxial stomatal density, but covariance between density and length is similar on both surfaces. (a) Data from 236 phylogenetically independent contrasts of change in log(stomatal length) ($x$-axis) and log(stomatal density) ($y$-axis) for abaxial (left panel) and adaxial (right panel) leaf surfaces. Each contrast is shown by black points and every contrast appears on both panels. Grey ellipses are the model-estimated 95\% covariance ellipses. The negative covariance is similar for both surfaces but the breadth in the $y$-direction is larger for adaxial traits, indicating greater evolutionary divergence in log(stomatal density). (b) Parameter estimates (points), 66\% (thick lines), and 95\% HPD intervals for estimates of trait (co)variance. Grey points and lines represent ab- and adaxial values; black points and lines represent the estimated difference in (co)variance between surfaces. Only the variance for stomatal density (middle panel) is significantly greater for the adaxial surface (95\% HPD interval does not overlap the dashed line at 0). Reported parameter estimates are conditioned on zero time since divergence between taxa (see \protect\hyperlink{results}{Results}).}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/h2-raw.pdf}
\caption{Stomatal density on each surface is less constrained (lower correlation) than stomatal length (higher correlation). Relationship between stomatal density and length on each leaf surface in a synthesis of amphistomatous leaf traits across 638 taxa. The panels show the relationship between the abaxial trait value ($x$-axis) and the adaxial trait value ($y$-axis) on a log-log scale for stomatal density (left) and stomatal length (right). The dashed line in across the middle is the 1:1 line for reference.}
\label{fig:h2_raw}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/h2.pdf}
\caption{Developmental integration in stomatal length is much stronger than stomatal density between the surfaces of amphistomatous leaves (a) Data from 236 phylogenetically independent contrasts of change in the abaxial trait value ($x$-axis) against change in the adaxial trait value ($y$-axis) for log(stomatal density) (left panel) and log(stomatal length) (right panel). Each contrast is shown by black points and every contrast appears on both panels. Dashed grey lines are 1:1 lines for reference. Solid grey lines and ribbon the fitted SMA slope and 95\% HPD interval. (b) The SMA slope (left panel) is significantly less than 1 (isometry, top dashed line) for density but very close to isometric for length. The coefficient of determination ($r^2$, right panel) is also much greater for length than density. The points are parameter estimates with 66\% (thick lines) and 95\% HPD intervals. Reported parameter estimates are conditioned on zero time since divergence between taxa (see \protect\hyperlink{results}{Results}).}
\label{fig:h2}
\end{figure}

@@

@@

The final data set and phylogeny used in the analysis are included in the Online Appendix. The raw anatomical data and source code are deposited on Dryad (\url{https://datadryad.org/stash/share/Tk-7ZGwhdxyxAJu2dTsBdWDaId9qmog2fGYLWgGzPwU}.)

@@

\hypertarget{online-appendix}{%
\section{Online Appendix}\label{online-appendix}}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% \documentclass[11pt]{article} % redundant
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\usepackage{multirow} % for merging rows in tables
\usepackage{caption} % for \captionsetup

@@

\setstretch{1}
\begin{flushleft}
{\Large
\textbf\newline{How important are functional and developmental constraints on phenotypic evolution? An empirical test with the stomatal anatomy of flowering plants}
}
\newline
\\
Christopher D. Muir\textsuperscript{1,*} (ORCID: 0000-0003-2555-3878),
Miquel \`{A}ngel Conesa\textsuperscript{2},
Jeroni Galm\'{e}s\textsuperscript{2} (ORCID: 0000-0002-7299-9349),
Varsha S. Pathare\textsuperscript{3} (ORCID: 0000-0001-6220-7531),
Patricia Rivera\textsuperscript{4},
Rosana López Rodríguez\textsuperscript{5},
Teresa Terrazas\textsuperscript{4},
Dongliang Xiong\textsuperscript{6} (ORCID: 0000-0002-6332-2627)
\\
\bigskip
\bf{1} School of Life Sciences, University of Hawaii at M\=anoa, Honolulu, HI 96822, USA \\
\bf{2} Research Group on Plant Biology under Mediterranean Conditions, Departament de Biologia, Universitat de les Illes Balears, Ctra. Valldemossa km 7.5, E-07122, Palma, Spain \\
\bf{3} School of Biological Sciences, Washington State University, Pullman, WA 99164-4236, USA \\
\bf{4} Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Apartado Postal 70‑367, 04510 Mexico City, Mexico \\
\bf{5} Departamento de Sistemas y Recursos Naturales, Universidad Politécnica de Madrid, 28040 Madrid, Spain \\
\bf{6} National Key Laboratory of Crop Genetic Improvement, MOA Key Laboratory of Crop Ecophysiology and Farming System in the Middle Reaches of the Yangtze River, College of Plant Science and Technology, Huazhong Agricultural University, Wuhan, Hubei 430070, China\\
\bigskip
$\ast$ cdmuir@hawaii.edu

\textit{Keywords}: adaptation, amphistomy, developmental integration, leaf, packing limits, phylogenetic comparative methods, stomata

@@

\newpage{}

@@

Given the significance of stomata for plant function and global vegetation modeling (\protect\hyperlink{ref-berry_stomata:_2010}{Berry, Beerling, and Franks 2010}), we would like to understand what factors constrain their anatomical variation. Here we focus on interspecific variation in mean trait values rather than intraspecific variation. Packing constraints and developmental integration could explain inverse size-density scaling. The density and size of stomata on a planar leaf surface can be viewed as a packing problem where the total area allocated to stomata cannot exceed the total leaf area. This is a functional, or geometric, constraint because certain combinations of large size and high density are not physically possible. This functional constraint cannot explain why combinations of low density and small size are rare, but may explain why stomatal size must decrease when density increases as the leaf runs out of space. The packing limit of functional stomata is less than the entire leaf area, but the exact value is unclear. The realized upper limit is close to 1/3 or 1/2 (\protect\hyperlink{ref-de_boer_optimal_2016}{de Boer et al. 2016}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}; \protect\hyperlink{ref-liu_scaling_2021}{Liu et al. 2021}) for the species' mean, not an individual leaf.

@@

The second hypothesis is that cell size mediates developmental integration between stomatal size and density. If developmental integration is the primary reason for inverse size-density scaling, then amphistomatous leaves will exhibit identical size-density scaling on each surface. If the stomatal size and density scale differently on each surface, this implies that they can evolve independently and that selection against generally maladaptive trait combinations shapes their covariance. Furthermore, we predicted that divergence in genome size, which is strongly associated with meristematic cell volume (\protect\hyperlink{ref-simova_geometrical_2012}{Šímová and Herben 2012}), would covary with stomatal size and density similarly on each surface.

@@

Unless otherwise mentioned, we performed all data wrangling and statistical analyses in \emph{R} version 4.2.1 (\protect\hyperlink{ref-r_core_team_r:_2022}{R Core Team 2022}). Source code is publicly available on GitHub (\url{https://github.com/cdmuir/stomata-independence}) and archived on Zenodo (see Data availability).

@@

Developmental integration could shape patterns of phenotypic macroevolution, but a major hindrance to progress is that verbal models do not make precise, quantitative predictions that distinguish it from alternatives. An advantage of testing developmental integration in stomata is that their development is well studied (\protect\hyperlink{ref-bergmann_stomatal_2007}{Bergmann and Sack 2007}; \protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}). We can leverage that knowledge to build a developmental function and derive equations for phenotypic (co)variance caused by developmental integration. If observed patterns of evolution are inconsistent with developmental integration, theory may also help identify which parameters lead to developmental disintegration. We combined and extended stomatal development models to predict how stomatal density and length would diverge if stomatal development were constrained and how those predictions would change if stomatal development were unconstrained. We summarize our methods verbally here and direct readers interested in the mathematical details to Notes S1. A graphical summary is provided in Fig. \ref{fig:developmental-integration}. We imposed constraint by assuming the stomatal developmental function is constrained. The developmental function maps cell size prior to differentiation onto stomatal size and density using two parameters. The first parameter describes how cell volume is apportioned between epidermal cells and guard cell meristemoids during asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}). The second parameter is stomatal index (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}), which is determined by amplifying and spacing divisions after asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}). When these parameters are fixed, divergence in stomatal size and density is determined by divergence in meristematic cell volume and expansion prior to asymmetric division. We relaxed this constraint by treating parameters of the developmental function as random variables that can diverge between species. We used random variable algebra to derive predicted (co)variance in divergence between stomatal length and density (see \protect\hyperlink{ref-lynch_genetics_1998}{Lynch and Walsh 1998} for key random variable algebra theorems).

@@

We searched the literature for studies that measured stomatal density and stomatal size, either guard cell length or stomatal pore length, for both abaxial and adaxial leaf surfaces. In other words, we did not include studies unless they reported separate density and size values for each surface. We did not record leaf angle because it is typically not reported, but we presume that for the vast majority of taxa that the abaxial is the lower surface and the adaxial is the upper surface. This is reversed in resupinate leaves, but to the best of our knowledge, our synthesis did not include resupinate leaves. None of the species with resupinate leaves listed by Chitwood et al. (\protect\hyperlink{ref-chitwood_conflict_2012}{2012}) are in our data set. We refer to guard cell length as stomatal length and converted stomatal pore length to stomatal length assuming guard cell length is twice pore length (\protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}). Table \ref{tab:traits} lists anatomical traits and symbols. Abaxial traits are subscripted with \(_{\textrm{ab}}\); adaxial traits are subscripted with \(_{\textrm{ad}}\)

@@

We generated an ultrametric, bifurcating phylogeny of 638 taxa by resolving and removing ambiguous taxonomic names, placing taxa on the GBOTB.extended mega-tree of seed plants (\protect\hyperlink{ref-smith_constructing_2018}{S. A. Smith and Brown 2018}; \protect\hyperlink{ref-zanne_three_2014}{Zanne et al. 2014}), and resolving polytomies using published sequence data. Divergence times in this phylogeny are based on extensive fossil calibration (see \protect\hyperlink{ref-smith_constructing_2018}{S. A. Smith and Brown 2018}; \protect\hyperlink{ref-magallon_metacalibrated_2015}{Magallón et al. 2015} for further detail). The complete methodology is described in the online supplement (Notes S2).

@@

All hypotheses make predictions about trait (co)variance matrices or parameters derived from them (see Notes S1 and subsections below). Within and among species covariation is a hallmark of developmental integration (\protect\hyperlink{ref-armbruster_multilevel_1988}{Armbruster 1988}), but other evolutionary processes also lead to covariance. Distinguishing between them requires deriving predictions and testing whether observed covariance is consistent with one hypothesis or another. We estimated the \(4 \times 4\) covariance matrix of phylogenetically independent contrasts between log-transformed values of \(\Delta \text{log}(D_\mathrm{ab})\), \(\Delta \text{log}(D_\mathrm{ad})\), \(\Delta \text{log}(L_\mathrm{ab})\), and \(\Delta \text{log}(L_\mathrm{ad})\) using a distributional multiresponse robust Bayesian approach. See Table \ref{tab:traits} for variable definitions. We denote variances as \(\text{Var}[\Delta \text{log(trait)}]\) and covariances as \(\text{Cov}[\Delta \text{log(trait}_1),\Delta \text{log(trait}_2)]\). We used a multivariate \(t\)-distribution rather than a Normal distribution because estimates using the former are more robust to exceptional trait values (\protect\hyperlink{ref-lange_robust_1989}{Lange, Little, and Taylor 1989}). We also estimated whether the variance in trait divergence increases with time. Under many trait evolution models (e.g.~Brownian motion), interspecific variance increases through time. To account for this, we included time since taxon-pair divergence as an explanatory variable affecting the trait covariance matrix.

@@

We asked how divergence in stomatal length and density would covary if the developmental function were constrained and compared it to their divergence when the developmental function can evolve. When the developmental function is constrained this means that allocation to guard cell meristemoids during asymmetric division and stomatal index are fixed (see Notes S1 for mathematical description). Under these assumptions, divergence in stomatal length and density is mediated entirely by divergence in meristematic cell volume and expansion prior to differentiation. Developmental integration is strong because divergence in density is perfectly negatively correlated with divergence in size. In contrast, stomatal length and density can diverge independently when the developmental function is not fixed. Divergence in asymmetric cell division affects stomatal size independently of density; divergence in stomatal index affects stomatal density independently of size. Divergence in the developmental function causes developmental disintegration because stomatal density and size can diverge independently. Developmental integration is minimal when asymmetric cell division and/or stomatal index diverge more than meristematic cell volume and expansion. The three main conclusions are that 1) developmental constraint leads to developmental integration; 2) different (co)variance in divergence of stomatal length and density on each surface implies the developmental function is not fixed; and 3) divergence in different components of the developmental function affect stomatal length and density differently. See Notes S1 and Table \ref{tab:predictions} for more a complete derivation and detailed predictions.

@@

@@

@@

@@

\hypertarget{statement-of-authorship}{%
\section{Statement of Authorship}\label{statement-of-authorship}}

CDM led conceptualization, data analysis, and writing the original draft. All authors collected data and assisted with review and editing.

@@

The final data set and phylogeny used in the analysis are included in the Online Supplement. The raw anatomical data and source code are deposited in the Dryad Digital Repository: \url{https://doi.org/10.5061/dryad.0rxwdbs42} (\protect\hyperlink{ref-muir_data_2022}{Muir et al. 2022}).

@@

\hypertarget{tables}{%
\section{Tables}\label{tables}}
